### 2020-05-18 probabilisticProgrammingMartingales.tex
A more succinct version of the 10? (schematic) terms in Rch of geometric distribution example

\section{Language definition}
The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{MakOP20b}. Types and terms are defined as follows, where $r$ is a real, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function and $\Gamma$ is an environment:
\begin{align*}
  & \text{types } A, B ::= \textsf{R}  \mid  A \to B \\
  & \text{values } v ::= \lambda x.s  \mid  \underline{r} \\
  & \text{terms } s, t ::= v  \mid  x  \mid  t_1 t_2  \mid  \underline{f}(s_1,\dots ,s_n)  \mid  \tY s  \mid  \tif{s < 0}{t_1}{t_2}  \mid  \tsample  \mid  \tscore(s)
\end{align*}
\begin{align*}
  \frac{}{\Gamma ; x:A \vdash x:A} \qquad
  \frac{\Gamma ; x:A \vdash s : B}{\Gamma \vdash \lambda x.s : A \to B} \qquad
  \frac{}{\underline{r} : \textsf{R}} \qquad
  \frac{\Gamma \vdash s:A \to B \quad \Gamma \vdash t : A}{\Gamma \vdash s t : B} \\ \\
  \frac{\Gamma \vdash s_1:\textsf{R} \dots \Gamma \vdash s_n:\textsf{R}}{\Gamma \vdash \underline{f}(s_1,\dots,s_n) : \textsf{R}} \ f : \mathbb{R}^n \to \mathbb{R} \qquad
  \frac{\Gamma \vdash s : (A \to B) \to (A \to B)}{\Gamma \vdash \tY s : (A \to B)} \\ \\
  \frac{\Gamma \vdash c : \textsf{R} \quad \Gamma \vdash s_1 : A \quad \Gamma \vdash s_2 : A}{\Gamma \vdash \tif{c < 0}{s_1}{s_2} : A} \qquad
  \frac{}{\Gamma \vdash \tsample : \textsf{R}} \qquad
  \frac{\Gamma \vdash s : \textsf{R}}{\Gamma \vdash \tscore (s) : \textsf{R}}
\end{align*}

@@

Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $P_1$ and $P_2$, can overlap in a few different ways. If $P_1$ is an initial segment \lo{(i.e.~prefix)?} of $P_2$, then $N_2$ is also a subterm of $N_1$, which is written as $P_1 \leq P_2$. If neither $P_1 \leq P_2$ nor $P_1 \geq P_2$, the positions are said to be disjoint. This is mostly relevant in that in any substitution at some position, the subterms at all other positions are unaffected.

With this notation, a more general reduction relation $\to$ can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
\lo{On first reading, I struggled to parse the following clauses.}
\changed[lo]{The binary relation $\to$ is defined by the following rules, each is conditional on a redex occuring at position $P$ in the term $M$ in question.}
\begin{align*}
  \text{if } M \mid P = (\lambda x.s) v,\ & M \to M[s[v/x]/P] \\
  \text{if } M \mid P = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/P] \\
  \text{if } M \mid P = \tY \lambda x. s,\ & M \to M[\lambda z. s[(\tY \lambda x. s)/x] z/P] \text{ where $z$ is not free in $s$}\\
  \text{if } M \mid P = \tif{\underline r < 0}{s_1}{s_2},\ & M \to M[s_1/P] \text{ where }r < 0 \\
  \text{if } M \mid P = \tif{\underline r < 0}{s_1}{s_2},\ & M \to M[s_2/P] \text{ where }r \geq 0 \\
  \text{if } M \mid P = \tsample & \text{ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$},\\ & M \to M[\underline r/P] \text{ where } r \in [0,1] \\
  \text{if } M \mid P = \tscore(\underline r),\ & M \to M[\underline r/P].
\end{align*}
\lo{The definition above is clear. But as a general rule, it is good to illustrate definitions -- especially if they are genuinely new -- with an example or two.}

@@

Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a reduction sequence starting at $M$, and the second of which is a position in the final term of the reduction sequence. Reduction sequences are used rather than reachable terms because if the same term is reached twice, different samples may be needed, but they will be discussed as though they were terms, by abuse of notation. \lo{Readers would benefit greatly from some examples of an $L_0(M)$-pair, and relations $\sim_p$ and $\sim_c$, here, before the formal definitions. Explain the underlying intuition.} The relation $\sim$ is defined as the union of the symmetric relations $\sim_p$ and $\sim_c$ where
\begin{enumerate}
    \item If $N$ reduces to $O$ with the redex at position $P$\lo{Do you mean: $M \to^\ast N \to O = N[\dotDelta / P]$ for some redex contraction $\Delta \to \dotDelta$}, and $Q$ is a position in $N$ 
    where $Q \not \geq P$, then $(N,Q) \sim_p (O,Q)$.
    \lo{This implies that $Q$ is a position in $O$, but $O \mid Q$ is not guaranteed to exist. E.g.~$N = (\lambda x.x) 1$ and $O = 1$. Now $N \mid Q$ exists, where $Q = @_1 ; \lambda$, but $O \mid Q$ does not exist.}
    \lo{Ah! I overlooked the condition $Q \not\geq P =
    {\cdot}$.}

    \lo{I find this confusing. Can you express it formally?}

    \item If $N$ $\beta$-reduces to $O$ at position $P$, $Q$ is a position in $N \mid P;@_1;\lambda$ and $N \mid P;@_1;\lambda;Q$ is not the variable involved in the reduction, $(N,P;@_1;\lambda;Q) \sim_p (O, P;Q)$
    \item If $N$ $\textsf{if}$-reduces to $O$ at position $P$, with the first resp. second branch being taken, and $P;\textsf{if}_i;Q$ occurs in $N$ (where $i = 2$ resp. $3$), $(N,P;\textsf{if}_i;Q) \sim_p (O,P;Q)$
    \item If $N \mid P = (\lambda x. A) B$, $A \mid Q = x$, $B \mid R = C$ is a redex reducing to $C'$ and $S$ exists in $C'$, then letting $O_1$ be $N$ reduced first at $P$ then at $P;Q;R$ and letting $O_2$ be $N$ reduced first at $P;@_2;R$ then at $P$, $(O_1,P;Q;R;S) \sim_c (O_2,P;Q;R;S)$
    \item If $N \mid P = (\tY \lambda x. A)$, $A \mid Q = C$ is a redex reducing to $C'$, $R$ exists in $C'$ and $C' \mid R \neq x$, then letting $O_1$ be $N$ reduced first at $P$ then $P;\lambda;@_1;Q$ and $O_2$ reduced at $P;\tY;\lambda;Q$ then \changed[lo]{at} $P$, $(O_1, P;\lambda;@_1;Q;R) \sim_c (O_2, P;\lambda;@_1;Q;R)$
    \item If $N \mid P = (\tY \lambda x. A)$, $A \mid Q = C$ is a redex reducing to $C'$, $R$ exists in $C'$ and $A \mid S = x$, then letting $O_1$ be $N$ reduced first at $P$ then $P;\lambda;@_1;S;\tY;\lambda;Q$ and $O_2$ reduced at $P;\tY;\lambda;Q$ then \changed[lo]{at} $P$, 
    \[
    (O_1, P;\lambda;@_1;S;\tY;\lambda;Q;R) \sim_c (O_2, P;\lambda;@_1;S;\tY;\lambda;Q;R)
    \]
\end{enumerate}

@@

\begin{example}
See \Cref{fig:adk-diagram} for examples of $\sim_p$- and $\sim_c$-rules.
\end{example}

@@

In summary, all equivalence classes under $\sim^*$ are of the shapes \akr{This would be far easier to show with a few pictures. I just don't yet know how to make them in Latex and I'm prioritising getting the text done.}
\lo{I can suggest possible \LaTeX\ packages for drawing. Can you draw them and upload a scan?}

@@

\subsection{Partial ranking functions}
Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it. Take, for example, the term
\begin{align*}
&(\tY \lambda f, n: \\
&\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
&) \underline{0}
\end{align*}
which generates a geometric distribution.
    Despite its simplicity, its $Rch$ contains all the terms
\begin{itemize}
    \item $(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{0}$
    \item $(\lambda z.(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) z) \underline{i}$
    \item $(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{i}$
    \item $(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{n}{f (m+1)}) (m+1)}) \underline{i}$
    \item $\tif{\tsample - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
    \item $\tif{\underline r - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
    \item $\tif{\underline{r - 0.5} < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
    \item $\underline{i}$
    \item $(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)$
    \item $(\lambda z.(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) z) (\underline{i} + 1)$.
\end{itemize}
\akr{TODO: make this into a diagram with arrows.}
Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as taking only 3 distinct values.

>>>>

\section{Language definition}
The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{MakOP20b}. Types and terms are defined as follows, where $r$ is a real, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function and $\Gamma$ is an environment:
\begin{align*}
  & \text{types } A, B ::= \textsf{R}  \mid  A \to B \\
  & \text{values } v ::= \lambda x.s  \mid  \underline{r} \\
  & \text{terms } s, t ::= v  \mid  x  \mid  t_1 t_2  \mid  \underline{f}(s_1,\dots ,s_n)  \mid  \tY s  \mid  \tif{s < 0}{t_1}{t_2}  \mid  \tsample  \mid  \tscore(s)
\end{align*}
\begin{align*}
  \frac{}{\Gamma ; x:A \vdash x:A} \qquad
  \frac{\Gamma ; x:A \vdash s : B}{\Gamma \vdash \lambda x.s : A \to B} \qquad
  \frac{}{\underline{r} : \textsf{R}} \qquad
  \frac{\Gamma \vdash s:A \to B \quad \Gamma \vdash t : A}{\Gamma \vdash s \, t : B} \\ \\
  \frac{\Gamma \vdash s_1:\textsf{R} \dots \Gamma \vdash s_n:\textsf{R}}{\Gamma \vdash \underline{f}(s_1,\dots,s_n) : \textsf{R}} \ f : \mathbb{R}^n \to \mathbb{R} \qquad
  \frac{\Gamma \vdash s : (A \to B) \to (A \to B)}{\Gamma \vdash \tY s : (A \to B)} \\ \\
  \frac{\Gamma \vdash c : \textsf{R} \quad \Gamma \vdash s_1 : A \quad \Gamma \vdash s_2 : A}{\Gamma \vdash \tif{c < 0}{s_1}{s_2} : A} \qquad
  \frac{}{\Gamma \vdash \tsample : \textsf{R}} \qquad
  \frac{\Gamma \vdash s : \textsf{R}}{\Gamma \vdash \tscore (s) : \textsf{R}}
\end{align*}

@@

Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $P_1$ and $P_2$, can overlap in a few different ways. If $P_1$ is an initial segment (i.e.~prefix) of $P_2$, then $N_2$ is also a subterm of $N_1$, which is written as $P_1 \leq P_2$. If neither $P_1 \leq P_2$ nor $P_1 \geq P_2$, the positions are said to be disjoint. This is mostly relevant in that in any substitution at some position, the subterms at all other positions are unaffected.

With this notation, a more general reduction relation $\to$ can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
%\lo{On first reading, I struggled to parse the following clauses.}
\changed[lo]{The binary relation $\to$ is defined by the following rules, each is conditional on a redex occuring at position $P$ in the term $M$.}
\begin{align*}
  \text{if } M \mid P = (\lambda x.s) v,\ & M \to M[s[v/x]/P] \\
  \text{if } M \mid P = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/P] \\
  \text{if } M \mid P = \tY \lambda x. s,\ & M \to M[\lambda z. s[(\tY \lambda x. s)/x] z/P] \text{ where $z$ is not free in $s$}\\
  \text{if } M \mid P = \tif{\underline r < 0}{s_1}{s_2},\ & M \to M[s_1/P] \text{ where }r < 0 \\
  \text{if } M \mid P = \tif{\underline r < 0}{s_1}{s_2},\ & M \to M[s_2/P] \text{ where }r \geq 0 \\
  \text{if } M \mid P = \tsample & \text{ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$},\\ & M \to M[\underline r/P] \text{ where } r \in [0,1] \\
  \text{if } M \mid P = \tscore(\underline r),\ & M \to M[\underline r/P].
\end{align*}
\lo{The definition above is clear. But as a general rule, it is good to illustrate definitions -- especially if they are genuinely new -- with an example or two.}

@@

Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a reduction sequence starting at $M$, and the second of which is a position in the final term of the reduction sequence. Reduction sequences are used rather than reachable terms because if the same term is reached twice, different samples may be needed, but they will be discussed as though they were terms, by abuse of notation. \lo{Readers would benefit greatly from some examples of an $L_0(M)$-pair, and the relations $\sim_p$ and $\sim_c$, here, before the formal definitions. Explain the underlying intuition.} The relation $\sim$ is defined as the union of the symmetric relations $\sim_p$ \changed[lo]{($p$ for parent-child relationship)} and $\sim_c$ \changed[lo]{($c$ for cousin relationship)} where
\lo{As defined, $\sim_p$ is not a symmetric relation.}
\begin{enumerate}
    \item If $N$ reduces to $O$ with the redex at position $P$\lo{Do you mean: $M \to^\ast N \to O = N[\dotDelta / P]$ for some redex contraction $\Delta \to \dotDelta$}, and $Q$ is a position in $N$ 
    where $Q \not \geq P$, then $(N,Q) \sim_p (O,Q)$.
    %\lo{This implies that $Q$ is a position in $O$, but $O \mid Q$ is not guaranteed to exist. E.g.~$N = (\lambda x.x) 1$ and $O = 1$. Now $N \mid Q$ exists, where $Q = @_1 ; \lambda$, but $O \mid Q$ does not exist.} \lo{Ah! I overlooked the condition $Q \not\geq P = {\cdot}$.}

    \item If $N$ $\beta$-reduces to $O$ at position $P$, $Q$ is a position in $N \mid P;@_1;\lambda$ and $N \mid P;@_1;\lambda;Q$ is not the variable involved in the reduction, $(N,P;@_1;\lambda;Q) \sim_p (O, P;Q)$

    \item If $N$ $\textsf{if}$-reduces to $O$ at position $P$, with the first resp. second branch being taken, and $P;\textsf{if}_i;Q$ occurs in $N$ (where $i = 2$ resp. $3$), $(N,P;\textsf{if}_i;Q) \sim_p (O,P;Q)$

    \item If $N \mid P = (\lambda x. A) B$, $A \mid Q = x$, $B \mid R = C$ is a redex reducing to $C'$ and $S$ exists in $C'$, then letting $O_1$ be $N$ reduced first at $P$ then at $P;Q;R$ and letting $O_2$ be $N$ reduced first at $P;@_2;R$ then at $P$, $(O_1,P;Q;R;S) \sim_c (O_2,P;Q;R;S)$

    \item If $N \mid P = (\tY \lambda x. A)$, $A \mid Q = C$ is a redex reducing to $C'$, $R$ exists in $C'$ and $C' \mid R \neq x$, then letting $O_1$ be $N$ reduced first at $P$ then $P;\lambda;@_1;Q$ and $O_2$ reduced at $P;\tY;\lambda;Q$ then \changed[lo]{at} $P$, $(O_1, P;\lambda;@_1;Q;R) \sim_c (O_2, P;\lambda;@_1;Q;R)$

    \item If $N \mid P = (\tY \lambda x. A)$, $A \mid Q = C$ is a redex reducing to $C'$, $R$ exists in $C'$ and $A \mid S = x$, then letting $O_1$ be $N$ reduced first at $P$ then $P;\lambda;@_1;S;\tY;\lambda;Q$ and $O_2$ reduced at $P;\tY;\lambda;Q$ then \changed[lo]{at} $P$, 
    \[
    (O_1, P;\lambda;@_1;S;\tY;\lambda;Q;R) \sim_c (O_2, P;\lambda;@_1;S;\tY;\lambda;Q;R)
    \]
\end{enumerate}

@@

\begin{example}
See \Cref{fig:adk-diagram} for a graphical depiction of $\sim_p$- and $\sim_c$-rules.
\end{example}

@@

In summary, all equivalence classes under $\sim^*$ are of the shapes \akr{This would be far easier to show with a few pictures. I just don't yet know how to make them in Latex and I'm prioritising getting the text done.}
\lo{PGF/TikZ is the standard \LaTeX\ package for generating vector graphics. The manual \url{https://ctan.org/pkg/pgf?lang=en} is a 1318-page tome! There are numerous tutorials / guides on the Internet. E.g.~\url{https://www.overleaf.com/learn/latex/TikZ_package}}

@@

\subsection{Partial ranking functions}
Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it. Take, for example, the term
\begin{align*}
&(\tY \lambda f, n: \\
&\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
&) \underline{0}
\end{align*}
which generates a geometric distribution.
    Despite its simplicity, its $Rch$ contains all the terms
\begin{enumerate}
    \item $(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{0}$
    \item $(\lambda z.(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) z) \underline{i}$
    \item $(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{i}$
    \item $(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{n}{f (m+1)}) (m+1)}) \underline{i}$
    \item $\tif{\tsample - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
    \item $\tif{\underline r - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
    \item $\tif{\underline{r - 0.5} < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
    \item $\underline{i}$
    \item $(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)$
    \item $(\lambda z.(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) z) (\underline{i} + 1)$.
\end{enumerate}
\akr{TODO: make this into a diagram with arrows.}

\newpage

\lo{Here is a more readable version.}

Writing $\Theta = \lambda f, n : \tif{\tsample - 0.5 < 0}{n}{f (n+1)}$, we have
\begin{enumerate}
    \item $(\tY \, \Theta) \; \underline{0}$

    \item $\big(\lambda z.\Theta \; (\tY \, \Theta) \; z \big) \; \underline{i}$

    \item $\Theta \; (\tY \, \Theta) \; \underline{i}$

    \item \changed[akr]{$(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{n}{f (m+1)}) (m+1)}) \; \underline{i}$}

    \lo{I think the preceding should be: $(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (n+1)}) \; \underline{i}$, and so, more succinctly:
    \[
    \big(
    \lambda n: 
    \tif{\tsample - 0.5 < 0}{n}{(\tY \, \Theta) \; (n+1)} 
    \big) 
    \; \underline{i}
    \]}

    \item $\tif{\tsample - 0.5 < 0}{\underline{i}}{
    (\tY \, \Theta) \, (\underline{i}+1)}$

    \item $\tif{\underline r - 0.5 < 0}{\underline{i}}{(\tY \, \Theta) (\underline{i}+1)}$

    \item $\tif{\underline{r - 0.5} < 0}{\underline{i}}{(\tY \, \Theta) (\underline{i}+1)}$

    \item $\underline{i}$

    \item $(\tY \, \Theta) \; (\underline{i}+1)$

    \item $\big(\lambda z. \Theta \; (\tY \, \Theta) \;  z\big) \; (\underline{i} + 1)$.
\end{enumerate}

Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as taking only 3 distinct values.

### 2020-05-18 ranking-supermartingales.tex
ongoing
A more succinct version of the 10? (schematic) terms in Rch of geometric distribution example

\lo{We need to show that each $f(M_n)$ is integrable. LO claimed earlier that $0 \leq f(N) \leq f(M)$ for all $N \in \mathit{Rch}(M)$.
This is of course false: take $f$ with $v \, \tsample \mapsto 1$ and $v \, \underline{r} \mapsto 2\, r$ for some value $v$. 
However, it is true that
\[
\int_{S} \mu(\dif s) \, |f(M_n)(s)| 
=
\int_{S} \mu(\dif s) \, f(M_n)(s)
\leq
\int_{S} \mu(\dif s) \, f(M_0)(s)
=
f(M)
\]
the inequality above follows from \Cref{lem:key rankable}.
}

We say that a given SPCF term is \emph{type-1} (respectively 2, 3 and 4) if it has the shape $E[\tY (\lambda x. N)]$ (respectively $E[\tsample]$, $E[R]$ where $R$ is any other redex, and of a value).
Define $\mathbf{T}_i := 
\set{s \mid M_n(s) \hbox{ is type-$i$}}$.
It is straightforward to see that each $\mathbf{T}_i \in \calF_n$, and $\set{\mathbf{T}_1, \mathbf{T}_2, \mathbf{T}_3, \mathbf{T}_4}$ is a partition of $S$.
%By abuse of language, we refer to the three respective types of redexes as type-$1$, 2 and 3.
Hence it suffices to prove the following lemma.

@@

\medskip
As before, fix a $M \in \Lambda^0$.
Define random variables on $(S, \calF, \mu)$:
\begin{align*}
T_0(s) & := 0 \\
T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value or of form } E[\tY (\lambda x. N)] \}\\
X_n(s) & := f(M_n(s)) \\
Y_n(s) & := X_{T_n}(s) \\
T(s) &:= \min \set{n \mid Y_n(s) = 0}
\end{align*}
\lo{Define the \emph{$\tY$-runtime of $M$} to be the random variable $T^{\tY}_M$: 
\begin{itemize}
\item if $T_M(s) = n < \infty$ then $T^{\tY}_M(s) :={}$ number of $\tY$-reduction steps in the $n$ steps of reduction to $\red^n(M, s)$
\item if $T_M(s) = \infty$ then $T^{\tY}_M(s) := \infty$.
\end{itemize}
Observe that $T = T^{\tY}_M$.}
Note that 
\begin{enumerate}
\item $T_M < \infty$ a.s.~iff $T_M^{\tY} < \infty$ a.s.~(because $T_M(s) < \infty$ iff $T_M^{\tY}(s) < \infty$)
\item $\expect{T_M} <{} \infty$ implies $\expect{T_M^{\tY}} < \infty$ (because $T_M^{\tY} \leq T_M$), but the converse is false (\Cref{ex:tY finite does not imply t finite}).
\end{enumerate}
%Note that by definition $T_M^{\tY} \leq T_M$.

>>>>

\lo{We need to show that each $f(M_n)$ is integrable. LO claimed earlier that $0 \leq f(N) \leq f(M)$ for all $N \in \mathit{Rch}(M)$.
This is of false: take $f$ with $v \, \tsample \mapsto 1$ and $v \, \underline{r} \mapsto 2\, r$ for some value $v$. 
However, it is true that
\[
\int_{S} \mu(\dif s) \, |f(M_n)(s)| 
=
\int_{S} \mu(\dif s) \, f(M_n)(s)
\leq
\int_{S} \mu(\dif s) \, f(M_0)(s)
=
f(M)
\]
the inequality above follows from \Cref{lem:key rankable}.
}

We say that a given SPCF term is \emph{type-1} (respectively 2, 3 and 4) if it has the shape $E[\tY \lambda x. N]$ (respectively $E[\tsample]$, $E[R]$ where $R$ is any other redex, and of a value).
Define $\mathbf{T}_i := 
\set{s \mid M_n(s) \hbox{ is type-$i$}}$.
It is straightforward to see that each $\mathbf{T}_i \in \calF_n$, and $\set{\mathbf{T}_1, \mathbf{T}_2, \mathbf{T}_3, \mathbf{T}_4}$ is a partition of $S$.
%By abuse of language, we refer to the three respective types of redexes as type-$1$, 2 and 3.
Hence it suffices to prove the following lemma.

@@

\medskip
As before, fix a $M \in \Lambda^0$.
Define random variables on $(S, \calF, \mu)$:
\begin{align*}
T_0(s) & := 0 \\
T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value or of form } E[\tY \lambda x. N] \}\\
X_n(s) & := f(M_n(s)) \\
Y_n(s) & := X_{T_n}(s) \\
T(s) &:= \min \set{n \mid Y_n(s) = 0}
\end{align*}
\lo{Define the \emph{$\tY$-runtime of $M$} to be the random variable $T^{\tY}_M$: 
\begin{itemize}
\item if $T_M(s) = n < \infty$ then $T^{\tY}_M(s) :={}$ number of $\tY$-reduction steps in the $n$ steps of reduction to $\red^n(M, s)$
\item if $T_M(s) = \infty$ then $T^{\tY}_M(s) := \infty$.
\end{itemize}
Observe that $T = T^{\tY}_M$.}
Note that 
\begin{enumerate}
\item $T_M < \infty$ a.s.~iff $T_M^{\tY} < \infty$ a.s.~(because $T_M(s) < \infty$ iff $T_M^{\tY}(s) < \infty$)
\item $\expect{T_M} <{} \infty$ implies $\expect{T_M^{\tY}} < \infty$ (because $T_M^{\tY} \leq T_M$), but the converse is false (\Cref{ex:tY finite does not imply t finite}).
\end{enumerate}
%Note that by definition $T_M^{\tY} \leq T_M$.
